Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-08-11 main.tex
Use four panel figure for eigenvalues vs speed.
- Show uncontrolled/controlled & with/without rider for the 4 panels.
- Use the control law Marten included in his report to match what is on
  the bike.
Text edits and improvements.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
in which they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
\todo{Talk about the types of real life perturbations an how our perturbation
mimics it.}
%
\begin{figure}
  \centering
  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
the human controlled plant takes the form:

\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \vec{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7\todo{Double check
this is the right model}, Dieren, The Netherlands) with a custom motor in the
steering assembly capable of applying up to 7~\si{\newton\meter} of torque
between the head tube and steer tube, see
Figure~\ref{fig:balance-assist-bicycle}.  A custom motor controller converts
the commanded torque to applied torque. We measure the speed of the rear wheel
with an encoder (Magura XXX) and measure the body fixed roll rate of the
bicycle with a MEMs rate gyroscope (CRS 03-02s, Silicon Sensing, United
Kingdom).  \todo{Do you mean the gyro inside the bicycle or the external?
Currently filled in the data of the external one. Not sure what make and model
is inside the bicycle} The balance assist control algorithm is implemented on a
microprocessor (Teensy, PJRC, USA) and data from all sensors is logged with a
CAN bus (CanEdge2, CSS Electronics, Denmark) at approximately 100 Hz.
\todo{Sample rate is not the same for each sensor: wheel speed, imu = 1ms,
steer torque command is updated every 10ms, can log is written every 10ms}

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. The commanded steer torque
\(T^\textrm{m}_\delta\) from the steer motor follows the control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{weave} - v)\dot{\phi}
\end{align}
%
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid-rider dynamics. We use \(g=8\) and \(g=10\) as gain
values during the experiments. Scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

\begin{figure}
  \centering
  \subcaptionbox{}{
    \includegraphics[width=100mm]{figures/uncontrolled-eig-vs-speeds.png}
  }
  \subcaptionbox{}{
    \includegraphics[width=100mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
  }
  \caption{Uncontrolled (a) and controlled (b) root locus of eigenvalue
    components (real: solid, imaginary: dashed) plotted versus speed. Vertical
    blue lines indicate the two speeds we perform experiments at. Green shaded
    region is the linear stable range.}
  \label{fig:eigenvalues}
\end{figure}

\todo[inline]{Decide on what physical parameters to use.}

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino Mega 2560, Italy). We measure the force applied by each
motor at the handlebar via four inline load cells, each with a max load of
250~\si{\newton}. The left rear load cell  was damaged during pilot
experiments, and is replaced by a load cell with a max load of
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from 20 to 200~\si{\newton} per motor.  This results in a
torque between 16 and 160~\si{\newton\meter}. The four motors are arranged at
the four corners of a 1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group,
Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.  The
general design is described in detail in the Van De Velde's MSc
hesis~\citep{vandeVelde2022a} and the physical arrangement  is shown in
Figure~\ref{fig:participant-in-set-up}. Our modifications relative to that
described in the thesis include simplifying the controller with a inexpensive
microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
% NOTE : I get a compile error with the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

@@

The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $g=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred 2
weeks before the 6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
location of their front wheel on the center line of the treadmill. \todo{The
  results from this are not included because the balance-assist was turned on
for some participants} We define a ``fall'' on the treadmill by two criteria:
the rider removes their foot from the pedal and places it on the ground or the
bicycle wheel exceeds the width of the treadmill. We then applied perturbations
in random directions (clockwise or counter clockwise), starting at
20~\si{\newton} and increasing the magnitude by 30~\si{\newton\meter} until the
participants falls.  Figure~\ref{fig:perturbation-sequence} shows an example
result of a perturbation.  We log the magnitude that causes the first fall to
characterize that subject's nominal fall threshold. Following this we apply 20
perturbations of varying magnitudes to the cyclist while they ride at a
constant speed and record which perturbations cause a fall. We let the cyclist
rest and then perform another 20 perturbations. We randomize whether the
balance assist system is on or off during the first or second set of
perturbations.

%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. The
goal of this adaptive staircase procedure is to have participants fall for
approximately 50\% of the time. The procedure consists of twenty perturbations
per condition. Participants undergo two conditions: twenty perturbations with
the balance-assist turned on and twenty perturbations with the balance-assist
turned off.

Five possible perturbation forces are determined based on the initial
perturbation threshold estimate: the initial estimate itself, two perturbations
lower than the initial estimate and two perturbations higher than the initial
estimate.  The five possible perturbations are separated by 10 N steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased with 10~\si{\newton}, and vice versa
if the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied.

@@

Based on findings from measuring riders without balance
assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
may influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:

\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
\end{align}

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
in which they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
\todo{Talk about the types of real life perturbations an how our perturbation
mimics it.}
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
the human controlled plant takes the form:
%
\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \vec{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble Arroyo\todo{Double check this is the
right model}, Royal Dutch Gazelle, Dieren, The Netherlands) with a custom motor
in the steering assembly capable of applying up to 7~\si{\newton\meter} of
torque between the head tube and steer tube, see
Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller converts the
commanded torque to applied torque. We measure the speed of the rear wheel with
an encoder (Magura XXX) and measure the body fixed roll rate of the bicycle
with a MEMs rate gyroscope (\todo{check bike for this}). The balance assist
control algorithm is implemented on a microprocessor (Teensy, PJRC, USA) and
data from all sensors is logged with a CAN bus (CanEdge2, CSS Electronics,
Denmark) at at least 100 Hz. \todo{Sample rate is not the same for each sensor:
wheel speed, imu = 1ms, steer torque command is updated every 10ms, can log is
written every 10ms}

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}~\todo{Add the experimental weave mode
identificatio to this figure and consider lowering the gain on the model to
make the model match the experimental values.}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertia effect of a rigid rider (left) and with
    a rigid rider (right). Vertical blue lines indicate the two speeds we
    perform experiments at. Grey shaded region is the linear stable speed
    range. Geometric and inertial parameters for these plots were estimated
    with the methods presented in \cite{Moore2012} and software packages
    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in cooperation, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino Mega 2560, COMPANYXXX, Italy). We measure the force
applied by each motor at the handlebar via four inline 250~\si{\newton} load
cells. The left rear load cell was damaged during pilot experiments, and is
replaced by a load cell with a max load of 500~\si{\newton}. The commanded
force profiles are designed to apply an external pulsive torque to the front
assembly (handlebars, forks, wheel) at magnitudes varying from 16 to
160~\si{\newton\meter}. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group, Switzerland) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design of this system
is described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde include simplifying the controller
with a inexpensive microcontroller and the use of a simpler non-active safety
harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

@@

The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $g=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred
two weeks before the 6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, while attempting to
maintain the location of their front wheel on the center line of the treadmill.
We define a ``fall'' on the treadmill by two criteria: 1) the rider removes
their foot from the pedal and places it on the ground or 2) the bicycle wheel
exceeds the width of the treadmill belt. We then applied perturbations in
random directions (clockwise or counter-clockwise), starting at 20~\si{\newton}
and increasing the magnitude by 30~\si{\newton\meter} until the participants
fell. Figure~\ref{fig:perturbation-sequence} shows an example resulting motion
from a perturbation. We log the magnitude that causes the first fall to
characterize that subject's nominal fall threshold.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure~\cite{TODO} applying
perturbations above and below the initial perturbation threshold, while
allowing small progression of the perturbation threshold to accommodate
learning effects. Five possible perturbation forces are determined based on the
initial perturbation threshold estimate: the initial estimate itself, two
perturbations lower than the initial estimate and two perturbations higher than
the initial estimate.  The five possible perturbations are separated by 10 N
steps. For example, if the initial estimate of the perturbation threshold of a
participant is 80~\si{\newton}, the five possible perturbations are 60, 70, 80,
90 and 100~\si{\newton}. Which one of these five perturbations is chosen is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased with 10~\si{\newton}, and vice versa if
the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. We apply 20
perturbations in this manner while the subject rides at a constant speed and
record which perturbations cause a fall. We let the cyclist rest and then
perform another 20 perturbations. We randomize whether the balance assist
system is on or off during the first or second set of perturbations.

@@

Based on findings from measuring riders without balance
assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
may influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
\end{align}
